---
title: "Exercise #3"
subtitle: "Fortgeschrittene Statistische Software für NF"
author: "Dennis Gruenewald (12259635)"
date: "`r Sys.Date()`"
output: distill::distill_article
---

## General Remarks

-   You can submit your solutions in teams of up to 3 students.
-   Include all your team-member's names and student numbers
    (Matrikelnummern) in the `authors` field.
-   Please use the exercise template document to work on and submit your
    results.
-   Use a level 2 heading for each new exercise and answer each subtask
    next to its bullet points or use a new level 3 heading if you want.
-   Always render the R code for your solutions (`echo=TRUE`) and make
    sure to include the resulting data in your rendered document.
    -   Make sure to not print more than 10 rows of data (unless
        specifically instructed to).
-   Always submit both the rendered document(s) as well as your source
    Rmarkdown or Quarto document. Submit the files separately on moodle,
    **not** as a zip archive.
-   Submission format is HTML. Other formats will lead to a deduction of
    points.

## Exercise 1: Initializing git (4 Points)

For this whole exercise sheet we will be tracking all our changes to it
in git.

a)  Start by initializing a new R project with git support, called
    `2025-exeRcise-sheet-3`. If you forgot how to do this, you can
    follow this
    [guide](https://malikaihle.github.io/Introduction-RStudio-Git-GitHub/rstudio_project.html).
b)  Commit the files generated by Rstudio.
c)  For all of the following tasks in this exercise sheet we ask you to
    always commit your changes after finishing each subtask e.g. create
    a commit after task *1d*, *1e* etc.

> Note: This applies only to answers that have text or code as their
> answer. If you complete tasks in a different order or forget to commit
> one, this is no problem. If you change your answers you can just
> create multiple commits to track the changes.

d)  Name 2 strengths and 2 weaknesses of git. (Don't forget to create a
    commit after this answer, see *1c*)
    
Strengths: Backup in case of errors in code & create for working in groups or as a team
Weaknesses: Complicated in the beginning & slow with big data

e)  Knit this exercise sheet. Some new files will automatically be
    generated when knitting the sheet e.g. the HTML page. Ignore these
    files, as we only want to track the source files themselves. You
    can, but don't need to create a `.gitignore` file. Just do not
    commit these files manually.

## Exercise 2: Putting your Repository on GitHub (3 Points)

For this task you will upload your solution to GitHub.

a)  Create a new repository on GitHub in your account named
    `exeRcise-sheet-3`. Make sure you create a **public repository** so
    we are able to see it for grading. Add the link to the repository
    below:
    
https://github.com/Cardii1306/exeRcise-sheet-3

b)  Push your code to this new repository by copying and executing the
    snippet on github listed under
    `…or push an existing repository from the command line`.
c)  Regularly push your latest changes to GitHub again and especially do
    so when you are finished with this sheet.

## Exercise 3: Pixar Films (4 Points)

Download the `pixar_films` and `public_response` datasets from the
GitHub repository and track them in git.

Link:
<https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-03-11>

For small datasets like these adding them to git is not a problem.

a)  Load the `pixar_films` dataset into R. Clean the dataset by removing
    films without a title. Inspect the variable `film_rating`. What are
    the possible values and what do they mean? Create a factor variable
    for the film rating. Why is this appropriate?

```{r load pixarfilms and inspect, echo=TRUE}
library(tidyverse)
library(dplyr)

# read the pixar data
pixar_films <- read_csv("pixar_films.csv")

# have a look at the data
glimpse(pixar_films)

# remove na's to better work with the data
pixar_films_wo_na <- pixar_films %>%
  filter(!is.na(film))

glimpse(pixar_films_wo_na$film_rating)

# have a look at the unique film ratings
unique(pixar_films_wo_na$film_rating)
```


The possible values for "film_rating" are G, PG or N/A. G means that the film is for "all ages admitted" and PG means that the "Parental Guidance" is suggested. "N/A" means that there is no value available for a specific film. 

```{r Create factor variable from film_raing, echo=TRUE}
# create caractor vevtor from ratings
film_rating <- c("G", "PG")
factor_film_rating <- factor(film_rating, levels = film_rating, ordered = TRUE)
```

It is appropriate because the different film_ratings can be ordered. "G" is lower age than "PG" so factoring the variable makes it easiert to work with it. 


b)  Inspect the film titles manually. Which films form a film series? A
    film series can be identified by a common word in the titles of the
    films, often in conjunction with a number in the title,
    e.g. "Despicable Me" and "Despicable Me 2". Create a dataframe which
    displays a list of the different series with the titles of the films
    and how many films belong to the series. Output the dataframe.
    
```{r Inspect for series, echo=TRUE}
glimpse(pixar_films_wo_na)

# glimpse does not show enough, so i had to take a look in the dataframe
```

The Film Series are Cars, Finding Dori & Nemo, Monsters Movies, Incredibles & Toy Story. 

```{r create df with series, echo=TRUE}
# catch if the film is in a series and name them in an extra culumn after their series name
pixar_films_series <- pixar_films_wo_na %>%
  mutate(
    series = case_when(
      str_detect(film, "Cars") ~ "Cars",
      str_detect(film, "Finding") ~ "Finding Dori & Nemo",
      str_detect(film, "Monsters") ~ "Monsters",
      str_detect(film, "Incredibles") ~ "Incredibles",
      str_detect(film, "Toy Story") ~ "Toy Story",
    )
  )

# group the films
pixar_films_series_grouped <- pixar_films_series %>%
  filter(!is.na(series)) %>%
  group_by(series) %>%
  summarise(
    film_titles = paste (film, collapse = ", "),
    number_of_films = n(),
  )
```


c)  Load the `public_response` dataframe into R. Convert the
    `cinema_score` variable into a factor while ensuring the factor
    levels are defined in ascending order, from the lowest to the
    highest score. Combine `public_response` with the `pixar_films`
    dataset using an appropriate merge variable.

```{r load and convert public response data, echo=TRUE}
# read the data
public_response <- read_csv("public_response.csv")

# have a look at the data
glimpse(public_response)

#check for unique cinema scores 
unique(public_response$cinema_score)

# create character vector for cinema score
cinema_score_levels <- c("A-", "A", "A+")

# make cinema score to a factor variable
public_response <- public_response %>%
  mutate(
    cinema_score = factor(cinema_score, levels = cinema_score_levels, ordered = TRUE)
  )
```

```{r combining the data, echo=TRUE}
# combine the dataset 
pixar_combined <- pixar_films_wo_na %>%
  left_join(public_response, by = "film") %>%
  left_join(pixar_films_series %>% select(film, series), by = "film")
```


d)  Choose one of the variables representing the public response and
    create a bar plot for the films belonging to a series. Here are the
    details of the plot:

    -   The film series are represented on the x-axis.
    -   Your chosen public response variable is displayed on the y-axis.
    -   Each film in the series is represented as a separate bar. Bars
        are grouped by film under their respective series on the x-axis.
        Order the bars within a series according to the release date of
        the films.
    -   A title and axis labels for context.

    What do you notice when comparing the scores of the films in a
    series? Do you see any patterns?
    
```{r arrange for series and filter out films that are not in series, echo=TRUE}
# clean the data for films that are not in series
pixar_combined <- pixar_combined %>%
  filter(!is.na(series)) %>%
  arrange(series, release_date) %>%
  mutate(film = factor(film, levels = unique(film)))
```

```{r creating plot to see how the films are perfoming within a series, echo=TRUE}
library(ggplot2)

# create a ggplot and use "identity" as metric scale
ggplot(pixar_combined, aes(x = series, y = metacritic, fill = film)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(
    title = "Metacritic Score for Pixar Film Series",
    x = "Series",
    y = "Metacritic",
    fill = "Film Title"
  ) +
  theme_minimal()
```

You can clearly notice that within a Series the Score in this case "Metacritic" is getting lower and lower the longer the series is. That is a typical phenomen a lot of film series do have, the viewer is less happy with the sequels. 

## Exercise 4: Open Analysis (4 points)

This exercise is a bit more open-ended. You can choose any dataset from
[Our World in Data](https://ourworldindata.org/) and analyze it, while
determining the research question yourself.

a)  Go to <https://github.com/owid/owid-datasets/tree/master/datasets>
    and choose a dataset that interests you. You can have a look at
    <https://ourworldindata.org/> to gather some inspiration.
    
b)  Download the dataset and track it in git.

c)  Put the name / title of the dataset and a link to it below.

- Dataset Name: Alcohol consumption in USA since 1850 — NIAAA  
- Link: <https://github.com/owid/owid-datasets/blob/master/datasets/Alcohol%20consumption%20in%20USA%20since%201850%20%E2%80%94%20NIAAA/Alcohol%20consumption%20in%20USA%20since%201850%20%E2%80%94%20NIAAA.csv>


d)  Come up with a (research) question you want to answer with the data
    and briefly explain why you believe this is an interesting question
    within one sentence. It should be a question that can be answered
    with the dataset and using R.
    
Is there a noticeable change in Alcohol consumption in the 150+ years or any trends? And is there a shift in between the different types of alcohol (beer, wine & sporits)?

e)  Use R to answer your chosen question. Please limit your analysis to
    the functions and techniques we have covered so far in the course.
    You are **not expected** to use advanced statistical models or
    external packages which haven't been introduced.
    
    

```{r loading and converting the data, echo=TRUE}
alcohol_data <- read_csv("Alcohol_consumption_US_form_1850.csv")
glimpse(alcohol_data)

# rename the columns
alcohol_data <- alcohol_data %>%
  rename(
    Year = Year,
    Beer = `Beer (NIAAA)`,
    Wine = `Wine (NIAAA)`,
    Spirits = `Spirits (NIAAA)`,
    All = `All beverages (NIAAA)`
  )

```

```{r check if before 1950 or after 1950 there was more alcohol consumption, echo=TRUE}
# group the period and calculate the average
alcohol_data %>%
  mutate(period = if_else(Year < 1950, "Before 1950", "1950 and after")) %>%
  group_by(period) %>%
  summarise(avg_total = mean(All, na.rm = TRUE))
```


```{r check how the average consumption has changen over 20 year periods, echo=TRUE}
# create 20year long periods and have a quick look at the data to check for trends
alcohol_data %>%
  mutate(period = floor((Year - 10) / 20) * 20 + 10) %>%
  group_by(period) %>%
  summarise(
    avg_beer = mean(Beer, na.rm = TRUE),
    avg_wine = mean(Wine, na.rm = TRUE),
    avg_spirits = mean(Spirits, na.rm = TRUE)
  )
```

From the functions above we can answer my question the following way: The alcohol consumption after 1950 is higher than in the period from 1850 to 1950. So there is an upwards shift in alcohol consumption. Moreover there is a clear indicator that the consumption of wine and beer is growing over the years but the consumption of spirits is shrinking.


f)  Create a meaningful plot / figure with the dataset. Make sure to
    provide a figure caption (via the chunk options / Rmarkdown) and
    correctly label the figure.
    
```{r plotting the data to check for trends, echo=TRUE, fig.cap="Figure 1: Alcohol consumption by type in the USA, 1850–present. Source: Our World in Data"}
# plot the conumption of alcohol on a line plot
ggplot(alcohol_data, aes(x = Year)) +
  geom_line(aes(y = Beer, color = "Beer")) +
  geom_line(aes(y = Wine, color = "Wine")) +
  geom_line(aes(y = Spirits, color = "Spirits")) +
  labs(
    title = "Alcohol Consumption Trends in the USA (1850–Present)",
    x = "Year",
    y = "Liters of Alcohol per Capita",
    color = "Alcohol Type",
    caption = "Source: Our World in Data (https://github.com/owid/owid-datasets/tree/master/datasets)"
  ) +
  theme_minimal()

```
The Figure supports what I said above, but there is one interessing fact to see. The Alcohol Consumption in the 1920s & 30s is at 0, that comes from the prohibition of alcohol in the United States. That could lead to a missinterpreation of the "before and after 1950" average alcohol consumption. It may be a good year to check if the alcohol conumption was higher or lower before or after the prohibition. So i will check for this as an addon. 

```{r check if before prohibition or after there was more alcohol consumption, echo=TRUE}
# group the data but "remove" the prohibition
alcohol_data %>%
  filter(Year < 1920 | Year > 1933) %>%
  mutate(period = if_else(Year < 1920, "Before Prohibition", "After Prohibition")) %>%
  group_by(period) %>%
  summarise(avg_total = mean(All, na.rm = TRUE))

```

It is interesting to see that the average alcohol consum is still higher after prohibition, but the difference is way smaller. So there is not a real effect to detect trough the prohibition. 


## Final Note

Make sure to push all your commits and changes to GitHub before
submitting the exercise sheet.
